Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-23 manuscript/response_to_reviews.Rmd

**Unless the reviewers and editor strongly disagree, we would request to keep this figure to clarify the main manuscript. We have revised the manuscript such that this figure is better integrated with the hypothesis testing (Table 1) and interpretation of results. We agree that similar patterns are common to any forest, and are not trying to claim this as any kind of novel result, but we're not aware of studies showing this depiction of height profiles for a temperate forest. (Such patterns are not shown in some relevant textbooks: Gordan Bonan's 2016 *Ecological Climatology* and Norman & Campbell's *Environmental Biophysics*. If we're missing studies that show such vertical profiles, we'd appreciate the references.)**

**We did create two new figures showing the results of the tree-ring analysis (and moved two tables to the SI).**

- Line 330-331: any evidence to support this? It looks rather speculative. As suggested above, does any species particularly show positive Rt or positive Rt are randomly distributed in your trees?

**We have removed the claim that increased growth may be due to reduced leaf area.**

@@

L67: I suggest toning down the sentence to something like ‘Wider xylem conduits are associated to higher vulnerability to embolism during drought’, as the mechanism linking wider conduits with embolism vulnerability is far from clear and probably indirect.

**We have revised this to read, "Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018]..."**

L81: you seem to use drought resistance and drought tolerance interchangeably, which may be confusing. I think it is important to clarify these definitions early on in the paper.

**We now define drought resistance in the first paragraph: " Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought,...". We then use this term when we wish to speak specifically about resistance. We use "tolerance" as a more general term, consistent with what seems to be a fairly nebulous definition in the literature.**

### 2020-07-23 manuscript/climate_sensitivity_draft.Rmd
clarify drought tolerance vs resistance
#124

Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the forest-dominated terrestrial carbon sink will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015]. Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015]. Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, and the extent to which their influence is consistent across droughts. 
Because the resistance and resilience of growth to drought is linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.

@@

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **3**) and associated traits at a single site [see also @elliott_forest_2015]. 
Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought tolerance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring-porous species are restricted mainly to temperate deciduous forests [@wheeler_variations_2007], while highly drought-tolerant diffuse-porous species exist in other biomes (**REFS**). 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
A second explanation of why *F. grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig. **S4**).
The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig. **2**).
Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.

@@

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. 
Our results, consistent with other observations around the world, imply that the tallest, most exposed trees will be most affected [@bennett_larger_2015; @stovall_tree_2019]. 
We show that, at least within the mature forest studied here, the vulnerability conferred by tall height and associated crown exposure outweigh any advantage of a larger root system, even in drier microenvironments. 
This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than below-ground environment. 
The same may not be true of systems where short trees exist outside of a buffered understory environment--*i.e.*, open grown trees or short-statured, early-successional forests.
The latter appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits. 
The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought tolerance and resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment. 
Further research on how leaf drought tolerance traits and drought vulnerability change over the course of succession would be valuable for addressing how drought tolerance changes as forests age [e.g. @rodriguez-caton_long-term_2015]. 
In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.

>>>>

Forests play a critical global role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the forest-dominated terrestrial carbon sink will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014; @dai_climate_2018], often despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015;  @cook_unprecedented_2015]. Droughts, intensified by climate change, have been affecting forests worldwide and are expected to continue as one of the most important drivers of forest change in the future [@allen_global_2010; @allen_underestimation_2015]. Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought, and the extent to which their influence is consistent across droughts. 
Because the resistance and resilience of growth to drought is linked to trees' probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality.
However, it has proven difficult to resolve the many factors affecting tree growth during drought with available forest census data, which only rarely captures extreme drought, and with tree-ring records, which capture multiple droughts but usually only sample a subset of a forest community, typically focusing on a single species or the largest individuals.

@@

The development of tree-ring chronologies for the twelve most dominant tree species at our site [@helcoski_growing_2019; @bourg_initial_2013] gave us the sample size to compare historical drought responses across species (Fig. **3**) and associated traits at a single site [see also @elliott_forest_2015]. 
Our study reinforced current understanding (see Introduction) that wood density and $LMA$ are not reliably linked to drought resistance (Table 1).
Contrary to previous studies in temperate deciduous forests, we did not find an association between xylem porosity and drought resistance, as the two diffuse-porous species, *Liriodendron tulipifera* and *Fagus grandifolia*, were at opposite ends of the $Rt$ spectrum (Fig. **3**). 
While the low $Rt$ of *L. tulipifera* is consistent with other studies [@elliott_forest_2015], the high $Rt$ of *F. grandifolia* contrasts with studies identifying diffuse porous species in general [@elliott_forest_2015; @kannenberg_linking_2019], and the genus *Fagus* in particular [@friedrichs_species-specific_2009], as drought sensitive.
There are two potential explanations for this discrepancy.
First, other traits can and do override the influence of xylem porosity on drought resistance. 
Ring-porous species are restricted mainly to temperate deciduous forests, while highly drought-tolerant diffuse-porous species exist in other biomes [@wheeler_variations_2007]. 
*Fagus grandifolia* had intermediate $\pi_{tlp}$ and low $PLA_{dry}$ (Fig. **S4**), which would have contributed to its drought resistance (Fig. **4**; see discussion below). 
A second explanation of why *F. grandifolia* trees at this particular site had higher $Rt$ is that the sampled individuals, reflective of the population within the plot, are generally shorter and in less-dominant canopy positions compared to most other species (Fig. **S4**).
The species, which is highly shade-tolerant, also has deep crowns [@andersonteixeira_size-related_2015], implying that a lower proportion of leaves would be affected by harsher microclimatic conditions at the top of the canopy under drought (Fig. **2**).
Thus, the high $Rt$ of the sampled *F. grandifolia* population can be explained by a combination of fairly drought-resistant leaf traits, shorter stature, and a buffered microenvironment.

@@

As climate change drives increasing drought in many of the world's forests [@trenberth_global_2014; @intergovernmental_panel_on_climate_change_climate_2015], the fate of forests and their climate feedbacks will be shaped by the biophysical and physiological drivers observed here. 
Our results, consistent with other observations around the world, imply that the tallest, most exposed trees will be most affected [@bennett_larger_2015; @stovall_tree_2019]. 
We show that, at least within the mature forest studied here, the vulnerability conferred by tall height and associated crown exposure outweigh any advantage of a larger root system, even in drier microenvironments. 
This would suggest that the drought responses of trees in mature forests are more strongly differentiated along the size spectrum by their above- than below-ground environment. 
The same may not be true of systems where short trees exist outside of a buffered understory environment--*i.e.*, open grown trees or short-statured, early-successional forests.
The latter appear to be limited more strongly by root water access during drought [@bretfeld_plant_2018], and would also be dominated by species with different traits. 
The earlier-successional species at our site (*Liriodendron tulipifera*, *Quercus spp.*, *Fraxinus americana*) display a mix of traits conferring drought resistance (Table 2), while the late-successional *Fagus grandifolia* displayed high drought resistance, in part because it exists primarily within a buffered microenvironment. 
Further research on how leaf drought tolerance traits and drought vulnerability change over the course of succession would be valuable for addressing how drought resistance changes as forests age [e.g. @rodriguez-caton_long-term_2015]. 
In the meantime, the results of this study advance our knowledge of the factors conferring drought resistance in a mature forest, opening the door for more accurate forecasting of forest responses to future drought.
